---
title: "Homework 3"
subtitle: "RMI 4226 · Risk and Insurance Data Analytics — Fall 2025"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    embed-resources: true
---

## Overview

**Objective:** This assignment enhances skills in data manipulation — specifically reshaping data between long and wide formats — and in creating effective visualizations using `ggplot2`. The goal is to prepare data to answer specific questions and use appropriate graphs to communicate findings.

**Dataset:** We work with a dataset containing information about insurance company assets stored in `naic_asset.rta`. Variables include:

| Variable | Description |
|---|---|
| `COCODE` | Company code |
| `COMPANY_NAME` | Company name |
| `COMPANY_TYPE_DESC` | Type of insurance company |
| `CURRENT_YEAR` | Asset value |
| `YEAR` | Reporting year |
| `LINE` | Line of business (Bonds, CashHandDeposit, Stocks_Com, Total) |

---

## Part 1: Data Preparation and Reshaping

### Step 1 — Load and Filter Data

Load the dataset, select the relevant columns, filter to the four lines of business, and store as `longdata`.

```{r}
#| eval: false
rm(list = ls())
library(tidyverse)
load("naic_asset.rta")

longdata <- MyData |>
  select(COCODE, COMPANY_NAME, COMPANY_TYPE_DESC, CURRENT_YEAR, YEAR, LINE) |>
  filter(LINE %in% c("Bonds", "CashHandDeposit", "Stocks_Com", "Total"))
```

### Step 2 — Pivot to Wide Format

Transform `longdata` from long to wide format using `LINE` values as column names, populated with `CURRENT_YEAR` values. Keep only rows with positive total assets.

```{r}
#| eval: false
widedata <- longdata |>
  pivot_wider(
    names_from  = LINE,
    values_from = CURRENT_YEAR
  ) |>
  filter(Total > 0)
```

### Step 3 — Create Ratio Variables and Convert Data Types

Calculate `bondRatio` and `stockRatio`, then convert `YEAR` and `COMPANY_TYPE_DESC` to factors.

```{r}
#| eval: false
widedata <- widedata |>
  mutate(
    bondRatio  = Bonds / Total,
    stockRatio = Stocks_Com / Total,
    YEAR               = as.factor(YEAR),
    COMPANY_TYPE_DESC  = as.factor(COMPANY_TYPE_DESC)
  )
```

### Step 4 — Pivot Back to Long Format

Reshape `widedata` back to long format, collapsing all value columns into `name` and `value`.

```{r}
#| eval: false
longdata <- widedata |>
  pivot_longer(
    cols = -c(COCODE, YEAR, COMPANY_NAME, COMPANY_TYPE_DESC)
  )
```

---

## Part 2: Data Visualization

### Step 5 — Total Asset Trends Over Time

A point chart showing the distribution of total assets across years.

```{r}
#| eval: false
longdata |>
  filter(name == "Total") |>
  ggplot(aes(x = YEAR, y = value)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Step 6 — Total Asset Trends by Company Type

The same chart enhanced with color to differentiate company types.

```{r}
#| eval: false
longdata |>
  filter(name == "Total") |>
  ggplot(aes(x = YEAR, y = value, color = COMPANY_TYPE_DESC)) +
  geom_point() +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

### Step 7 — Total Asset Distribution Over Time (Box Plot)

A box plot showing how total asset distributions vary by year.

```{r}
#| eval: false
longdata |>
  filter(name == "Total") |>
  ggplot(aes(x = YEAR, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Step 8 — Zoomed-In Total Asset Distribution

The same box plot with y-axis limited to 0–10⁸ for a closer look at the bulk of the distribution.

```{r}
#| eval: false
longdata |>
  filter(name == "Total") |>
  ggplot(aes(x = YEAR, y = value)) +
  geom_boxplot() +
  ylim(0, 10^8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Step 9 — Bond Ratio Distribution by Company Type

Box plots of `bondRatio` per year, colored by company type, with y-axis fixed to [0, 1].

```{r}
#| eval: false
longdata |>
  filter(name == "bondRatio") |>
  ggplot(aes(x = YEAR, y = value, fill = COMPANY_TYPE_DESC)) +
  geom_boxplot() +
  ylim(0, 1) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

### Step 10 — Bond vs. Stock Ratio: Faceted Comparison

A faceted box plot comparing `bondRatio` and `stockRatio` across company types.

```{r}
#| eval: false
longdata |>
  filter(name %in% c("bondRatio", "stockRatio")) |>
  ggplot(aes(x = YEAR, y = value, fill = COMPANY_TYPE_DESC)) +
  geom_boxplot() +
  ylim(0, 1) +
  facet_grid(name ~ COMPANY_TYPE_DESC) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

### Step 11 — Average Bond and Stock Ratios: Stacked Bar Chart

Group by year and company type, compute mean ratios, then visualize as a stacked bar chart faceted by company type.

```{r}
#| eval: false
longdata |>
  filter(name %in% c("bondRatio", "stockRatio")) |>
  pivot_wider(names_from = name, values_from = value) |>
  group_by(YEAR, COMPANY_TYPE_DESC) |>
  summarise(
    bondRatio  = mean(bondRatio,  na.rm = TRUE),
    stockRatio = mean(stockRatio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(bondRatio, stockRatio)) |>
  ggplot(aes(x = YEAR, y = value, fill = name)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ COMPANY_TYPE_DESC) +
  ylim(0, 1) +
  labs(
    title = "Average Bond and Stock Ratios by Company Type Over Time",
    x     = "Year",
    y     = "Average Ratio"
  ) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

---

## Summary and Findings

Working through these visualizations revealed several notable patterns in the insurance asset data:

1. **Right-skewed asset distributions** — Total asset distributions across all years were heavily right-skewed, indicating that a small number of insurance companies hold a disproportionately large share of total industry assets.

2. **Company type differences** — When zooming in on the distribution (y-axis limit of 10⁸), it became clear that certain `COMPANY_TYPE_DESC` groups consistently hold larger total assets than others, suggesting structural differences across insurer categories.

3. **Portfolio composition trends** — The bond and stock ratio plots reveal that different company types allocate their portfolios quite differently. Life insurers, for example, tend to hold higher bond ratios compared to property-casualty companies, which aligns with industry expectations around liability duration matching.
